A comparison of model-based and design-based approaches for spatial data.

### 2021-06-07 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
minor updates

We also typically assume that there is spatial correlation in $\bm{\delta}$, which can be modeled using a covariance function. Many common choices for this function assume that spatial covariance decreases with increasing Euclidean distance between sites. The primary function used throughout the simulations and applications of this manuscript is the Exponential covariance function: the $i,j^{th}$ entry for $\var(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = \theta_3 + \theta_1\exp(-h_{i,j}/\theta_2), \\
\textcolor{blue}{\sigma^2[(1 - v)\exp(-3\mathbf{h}_{i,j}/\phi) + v\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}]}, 
\end{align}
where $h_{i,j}$ is the distance between sites $i$ and $j$, and $\bm{\theta}$ is a vector of spatial covariance parameters of the partial sill $\theta_1$, the range $\theta_2$, and the nugget $\theta_3$. However, any spatial covariance function could be used in the place of the Exponential, including functions that allow for anisotropy [pg. 80 - 93][@chiles1999geostatistics].

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $\tau(\mathbf{b}'\mathbf{z})$ \textcolor{blue}{Did you mean to give the form of the BLUP here? $\tau(\mathbf{b}'\mathbf{z})$ is vague} and its prediction variance can be computed. While details of the derivation are in [@verhoef2008spatial], we note here that the predictor and its variance are both moment-based. Neither require a particular distribution for $\mathbf{z}$.

>>>>

We also typically assume that there is spatial correlation in $\bm{\delta}$, which can be modeled using a covariance function. Many common choices for this function assume that spatial covariance decreases with increasing Euclidean distance between sites. The primary function used throughout the simulations and applications of this manuscript is the Exponential covariance function: the $i,j^{th}$ entry for $\var(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align}
where $h_{i,j}$ is the distance between sites $i$ and $j$, and $\bm{\theta}$ is a vector of spatial covariance parameters of the partial sill $\theta_1$, the range $\theta_2$, and the nugget $\theta_3$. However, any spatial covariance function could be used in the place of the Exponential, including functions that allow for anisotropy [pg. 80 - 93][@chiles1999geostatistics].

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $\tau(\mathbf{b}'\mathbf{z})$ $\tau(\mathbf{b}'\mathbf{z})$ is vague} and its prediction variance can be computed. While details of the derivation are in [@verhoef2008spatial], we note here that the predictor and its variance are both moment-based. Neither require a particular distribution for $\mathbf{z}$.

### 2021-06-07 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
minor updates

Spatially balanced sampling algorithms use spatial information to obtain
samples spread out in space. Spatially balanced samples are useful
because they tend to yield estimators that are more precise than
estimators constructed from an sampling algorithm that is not spatially
balanced (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens Jr and
Olsen, 2004; Wang et al., 2013). To quantify spatial balance, Stevens Jr
and Olsen (2004) proposed statistics based on Voroni polygons. Many
spatially balanced sampling algorithms exist, including the Generalized
Random Tessellation Stratified (Stevens Jr and Olsen, 2004), the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
(Benedetti and Piersimoni, 2017), and Halton Iterative Partitioning
(Robertson et al., 2018) algorithms. Here we focus on the Generalized
Random Tessellation Stratified (GRTS) algorithm, which has several
attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample
frames. A finite sample frames contains a finite number of sampling
units and is related to a point geometry. An infinite sample frame
contains an infinite number of sampling units and is related to linear
and polygon geometries. Examples of point, linear, and polygon resources
include lake centroids, stream networks, and wetland areas,
respectively. In addition to its applicability for finite and infinite
sample frames, the GRTS algorithm naturally accommodates stratified
designs and designs with unequal selection probabilities. The algorithm
has also been used to select replacement sites using reverse
hierarchical ordering (Stevens Jr and Olsen, 2004). Replacement sites
are used to replace sites in the original sample that cannot be sampled,
often as a result of physical difficulty in reaching the site or
landowner denial of access to the sites. More recently, the GRTS
algorithm also accommodates legacy (historical) sites, minimum distance
between sites, and nearest neighbor replacement sites. The GRTS
algorithm is implemented in the \textbf{\textsf{R}} package
\texttt{spsurvey} (Dumelle et al., 2021).

@@

The Horvitz-Thompson estimator (Horvitz and Thompson, 1952) and its
continuous analog (Cordy, 1993) yield unbiased estimates of population
totals (and means). If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\), denoted by \(\hat{\tau}_{HT}\),
is given by \begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align} where \(Y_i\) is the observed value of the \(i\)th sampling
unit, and \(\pi_i\) is the inclusion probability of the \(i\)th sampling
unit. Estimating the variance of \(\hat{\tau}_{HT}\) directly relies on
knowing the probability that \(s_i\) and \(s_j\) are both included in
the sample (second-order inclusion probability), which can be unfeasible
to compute. Furthermore, this variance estimator does incorporate
spatial locations. To address these challenges, (Stevens Jr and Olsen,
2003) proposed an alternative variance estimator: the local neighborhood
variance estimator. The local neighborhood variance estimator does not
require second-order inclusion probabilities. It also incorporates
spatial locations, which tends to reduce the estimated variance of
compared \(\hat{\tau}_{HT}\) to a variance estimator ignoring spatial
location (Stevens Jr and Olsen, 2003).

@@

We also typically assume that there is spatial correlation in
\(\bm{\delta}\), which can be modeled using a covariance function. Many
common choices for this function assume that spatial covariance
decreases with increasing Euclidean distance between sites. The primary
function used throughout the simulations and applications of this
manuscript is the Exponential covariance function: the \(i,j^{th}\)
entry for \(\mathop{\mathrm{{var}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = \theta_3 + \theta_1\exp(-h_{i,j}/\theta_2), \\
\textcolor{blue}{\sigma^2[(1 - v)\exp(-3\mathbf{h}_{i,j}/\phi) + v\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}]}, 
\end{align} where \(h_{i,j}\) is the distance between sites \(i\) and
\(j\), and \(\bm{\theta}\) is a vector of spatial covariance parameters
of the partial sill \(\theta_1\), the range \(\theta_2\), and the nugget
\(\theta_3\). However, any spatial covariance function could be used in
the place of the Exponential, including functions that allow for
anisotropy {[}pg. 80 - 93{]}(Chiles and Delfiner, 1999).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(\tau(\mathbf{b}'\mathbf{z})\)
\textcolor{blue}{Did you mean to give the form of the BLUP here? $\tau(\mathbf{b}'\mathbf{z})$ is vague}
and its prediction variance can be computed. While details of the
derivation are in (Ver Hoef, 2008), we note here that the predictor and
its variance are both moment-based. Neither require a particular
distribution for \(\mathbf{z}\).

>>>>

Spatially balanced sampling algorithms use spatial information to obtain
samples spread out in space. Spatially balanced samples are useful
because they tend to yield estimators that are more precise than
estimators constructed from a sampling algorithm that is not spatially
balanced (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens Jr and
Olsen, 2004; Wang et al., 2013). To quantify spatial balance, Stevens Jr
and Olsen (2004) proposed statistics based on Voroni polygons. Many
spatially balanced sampling algorithms exist, including the Generalized
Random Tessellation Stratified (Stevens Jr and Olsen, 2004), the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
(Benedetti and Piersimoni, 2017), and Halton Iterative Partitioning
(Robertson et al., 2018) algorithms. Here we focus on the Generalized
Random Tessellation Stratified (GRTS) algorithm, which has several
attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample
frames. A finite sample frame contains a finite number of sampling units
and is related to a point geometry. An infinite sample frame contains an
infinite number of sampling units and is related to linear and polygon
geometries. Examples of point, linear, and polygon resources include
lake centroids, stream networks, and wetland areas, respectively. In
addition to its applicability for finite and infinite sample frames, the
GRTS algorithm naturally accommodates stratified designs and designs
with unequal selection probabilities. The algorithm has also been used
to select replacement sites using reverse hierarchical ordering (Stevens
Jr and Olsen, 2004). Replacement sites are used to replace sites in the
original sample that cannot be sampled, often as a result of physical
difficulty in reaching the site or landowner denial of access to the
sites. More recently, the GRTS algorithm also accommodates legacy
(historical) sites, minimum distance between sites, and nearest neighbor
replacement sites. The GRTS algorithm is implemented in the
\textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle et al., 2021).

@@

The Horvitz-Thompson estimator (Horvitz and Thompson, 1952) and its
continuous analog (Cordy, 1993) yield unbiased estimates of population
totals (and means). If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\), denoted by \(\hat{\tau}_{HT}\),
is given by \begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align} where \(Y_i\) is the observed value of the \(i\)th sampling
unit, and \(\pi_i\) is the inclusion probability of the \(i\)th sampling
unit. Estimating the variance of \(\hat{\tau}_{HT}\) directly relies on
knowing the probability that \(s_i\) and \(s_j\) are both included in
the sample (second-order inclusion probability), which can be unfeasible
to compute. Furthermore, this variance estimator does not incorporate
spatial locations. To address these challenges, (Stevens Jr and Olsen,
2003) proposed an alternative variance estimator: the local neighborhood
variance estimator. The local neighborhood variance estimator does not
require second-order inclusion probabilities. It also incorporates
spatial locations, which tends to reduce the estimated variance of
compared \(\hat{\tau}_{HT}\) to a variance estimator ignoring spatial
location (Stevens Jr and Olsen, 2003).

@@

We also typically assume that there is spatial correlation in
\(\bm{\delta}\), which can be modeled using a covariance function. Many
common choices for this function assume that spatial covariance
decreases with increasing Euclidean distance between sites. The primary
function used throughout the simulations and applications of this
manuscript is the Exponential covariance function: the \(i,j^{th}\)
entry for \(\mathop{\mathrm{{var}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align} where \(h_{i,j}\) is the distance between sites \(i\) and
\(j\), and \(\bm{\theta}\) is a vector of spatial covariance parameters
of the partial sill \(\theta_1\), the range \(\theta_2\), and the nugget
\(\theta_3\). However, any spatial covariance function could be used in
the place of the Exponential, including functions that allow for
anisotropy {[}pg. 80 - 93{]}(Chiles and Delfiner, 1999).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(\tau(\mathbf{b}'\mathbf{z})\)
\(\tau(\mathbf{b}'\mathbf{z})\) is vague\} and its prediction variance
can be computed. While details of the derivation are in (Ver Hoef,
2008), we note here that the predictor and its variance are both
moment-based. Neither require a particular distribution for
\(\mathbf{z}\).
